The Evolution of Phenotypically Invariant Genetic Networks

### 2017-10-23 kalman_walk_paper.tex
note on error in apx
minor latex corrections

\newcommand{\jss}[1]{{\color{olive}\it #1}}
% \newcommand{\ddt}{\frac{d}{dt}}
\newcommand{\ddt}{\dot}
\newcommand{\ro}{{ro}}
\newcommand{\nro}{{\bar{r}o}}
\newcommand{\rno}{{r\bar{o}}}
\newcommand{\nrno}{{\bar{r}\bar{o}}}
\newcommand{\reachable}{\mathcal{R}}
\newcommand{\unobservable}{\bar{\mathcal{O}}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\pda}{\frac{\partial}{\partial A_{ij}}}
\newcommand{\ind}{\mathds{1}}

@@

The description above does not completely describe how two diverging populations interact,
since the amount of \emph{segregation variance}, quantified by $R$,
will not stay constant.
To get an idea of how this might change,
suppose that a multivariate trait is determined by $L$ unlinked, biallelic loci,
and that the $i^\mathrm{th}$ locus has two alleles with additive effects $\pm x_i$,
so that begin homozygous for the $+$ allele contributes $+2x_i$ to the trait.
For simplicity, we will neglect the effects of selection.
\plri{fixup the below to be actually multivariate}
If the $+$ allele at locus $i$ is at frequency $p_i$ in a population,
then the mean and genetic variance of the trait in a diploid population with random mating is
\begin{align*}
    m &= 2 \sum_i p_i x_i \\
    s^2 &= 2 \sum_i p_i (1-p_i) x_i^2 .
\end{align*}

@@

Now consider an $F_2$ hybrid, where both parents are $F_1$
and so each heterozygous at locus $i$ with probability $p_i (1-q_i) + (1-p_i) q_i$.
Then
\begin{align*}
    R_2(p,q) &= \frac{1}{4} \sum_i \left(p_i (1-q_i) + (1-p_i) q_i \right) x_i^2  .
\end{align*}
Suppose that the two populations are slightly drifted from each other,
with frequency difference $p_i-q_i = 2\epsilon_i$.
Then,
$$
    p (1-q) + p (1-q)
    &=
    (u+\epsilon) (1-u+\epsilon) + (u-\epsilon) (1-u-\epsilon) \\
    &=
    2 u (1-u)
    % + \epsilon (u + (1-u) - u - (1-u))
    + 2 \epsilon^2 .
$$
If the frequencies have evolved neutrally in unconnected, Wright-Fisher populations 
of effective size $N$ for $t$ generations from a common ancestor with allele frequency $u$,
then $\epsilon$ has mean zero and variance roughly $u (1-u) t/N$.
Still assuming the populations are at stationarity,
so that $R_0$ is constant between the two,
and taking the frequencies $p_i$ as a proxy for the ancestral frequencies $u_i$,
this implies that we expect
\begin{align*}
    R_2 
    &\approx 
    R_0 + \frac{1}{2} \sum_i p_i (1-p_i) x_i^2 t/N \\
    &=
    \left( 1 + \frac{t}{N} \right) R_0 .
\end{align*}
On the other hand, the expected squared difference in trait \emph{means} here is
\begin{align} \label{eqn:var_dx}
    4 \sum_i p_i (1-p_i) x_i^2 t / N = 8 R_0 t /N .
\end{align}
This implies that under this model,
segregation variance in $F_2$s between two populations
is roughly increased by a factor of $1/8$ of the difference between their means.

>>>>

\newcommand{\jss}[1]{{\color{olive}\it #1}}
% \newcommand{\ddt}{\frac{d}{dt}}
\newcommand{\ddt}{\dot}
\newcommand{\ro}{{ro}}
\newcommand{\nro}{{\bar{r}o}}
\newcommand{\rno}{{r\bar{o}}}
\newcommand{\nrno}{{\bar{r}\bar{o}}}
\newcommand{\reachable}{\mathcal{R}}
\newcommand{\unobservable}{\bar{\mathcal{O}}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\var}{\mathop{\mbox{Var}}}
\newcommand{\cov}{\mathop{\mbox{Cov}}}
\newcommand{\pda}{\frac{\partial}{\partial A_{ij}}}
\newcommand{\ind}{\mathds{1}}

@@

The description above does not completely describe how two diverging populations interact,
since the amount of \emph{segregation variance}, quantified by $R$,
will not stay constant.
To get an idea of how this might change,
suppose that a multivariate trait is determined by $L$ unlinked, biallelic loci,
and that the $i^\mathrm{th}$ locus has two alleles with additive effects $\pm x_i$,
so that begin homozygous for the $+$ allele contributes $+2x_i$ to the trait.
For simplicity, we will neglect the effects of selection.
\plri{fixup the below to be actually multivariate}
If the $+$ allele at locus $i$ is at frequency $p_i$ in a population,
then the mean and genetic variance of the trait in a diploid population with random mating is
\plr{the mean should be $\sum_i x_i (2p_i-1)$}
\begin{align*}
    m &= 2 \sum_i p_i x_i \\
    s^2 &= 2 \sum_i p_i (1-p_i) x_i^2 .
\end{align*}

@@

Now consider an $F_2$ hybrid, where both parents are $F_1$
and so each heterozygous at locus $i$ with probability $p_i (1-q_i) + (1-p_i) q_i$.
Then
\begin{align*}
    R_2(p,q) &= \frac{1}{4} \sum_i \left(p_i (1-q_i) + (1-p_i) q_i \right) x_i^2  .
\end{align*}
Suppose that the two populations are slightly drifted from each other,
with frequency difference $p_i-q_i = 2\epsilon_i$.
Then,
\begin{align*}
    p (1-q) + p (1-q)
    &=
    (u+\epsilon) (1-u+\epsilon) + (u-\epsilon) (1-u-\epsilon) \\
    &=
    2 u (1-u)
    % + \epsilon (u + (1-u) - u - (1-u))
    + 2 \epsilon^2 .
\end{align*}
If the frequencies have evolved neutrally in unconnected, Wright-Fisher populations 
of effective size $N$ for $t$ generations from a common ancestor with allele frequency $u$,
then $\epsilon$ has mean zero and variance roughly $u (1-u) t/N$.
Still assuming the populations are at stationarity,
so that $R_0$ is constant between the two,
and taking the frequencies $p_i$ as a proxy for the ancestral frequencies $u_i$,
this implies that we expect
\begin{align*}
    R_2 
    &\approx 
    R_0 + \frac{1}{2} \sum_i p_i (1-p_i) x_i^2 t/N \\
    &=
    \left( 1 + \frac{t}{N} \right) R_0 .
\end{align*}
On the other hand, the expected squared difference in trait \emph{means} here is
\begin{align} \label{eqn:var_dx}
    4 \sum_i p_i (1-p_i) x_i^2 t / N = 8 R_0 t /N .
\end{align}
This implies that under this model,
segregation variance in $F_2$s between two populations
is roughly increased by a factor of $1/8$ of the difference between their means.
